---
title: "Introduction to Markdown and knitr"
author: "Alberto Roverato"
font: 12pt
output:
  html_document:
    toc: true
    # number_sections: true
  pdf_document:
    toc: true
    # number_sections: true
  word_document:
    toc: true
    # number_sections: true
---

\

## What is Markdown?

-   Markdown is a (simple) markup language for creating formatted text using a plain-text editor.

-   R Markdown is a free, open source tool that is installed like any other R package.

-   In fact, you need the *rmarkdown* package, but you don't need to explicitly install it or load it, as RStudio automatically does both when needed.

-   It enables you to keep all of your code, results, plots, and writing in one place.

-   It allows you to export your work to numerous formats including PDF, Microsoft Word, a slideshow, or an HTML document for use in a website.

-   Some examples are given below, but see also the *Markdown Quick Reference* in the *Help*

\

## 1. YAML header

YAML header is a short blob of text, specially formatted with `key: value` pairs tags, that seats at the top of an Rmarkdown document. It is demarcated by three dashes `---` on either end. Instances of YAML arguments are "title", "author", and "output".

\

## 2. Markdown syntax

### 2.1 Headers

This is how to obtain *headers* of different size

# Header level 1

## Header level 2

### Header level 3

#### Header level 4

##### Header level 5

###### Header level 6

\

### 2.2 Italic and Bold

This is *italic text*

This is also _italic text_

This is **bold text**

This is also __bold text__

And this is ***both bold and italic text***

And this is also ___both bold and italic text___

Markdown applications don't agree on how to handle underscores in the middle of a word. For compatibility, use asterisks to bold and italicize the middle of a word for emphasis.

Note that, as explained below, if you use the `RStudio’s visual markdown editor`, then underscores (\_) will be converted into asterisks (\*).

\

### 2.3 Highlighted text and rules

You can also highlight a text using a pair of backticks 
so that \`like this\` will be printed as `like this`

An horizontal rule can be obtained with `***`

***

but also with `---`

---

or with `___`

___

\

### 2.4 Ordered lists

These is how to do an ordered list:

1.  the first item

2.  the second item

3.  the third item

    -   one unordeded sub-item
    -   one unordered sub-item

\

This is an unordered list:

-   the first unordered item

-   the second unordered item

-   the third unordered item

    -   the first unordered sub-item
    -   the second unordered sub-item

\

### 2.5 Backslash

There are multiple ways to add a blank line in markdown. One way is to use the backslash (`\`) symbol. For example you can add two blank lines...

\
\

...like this.

Another use of backslash escapes is to generate literal characters which would otherwise have special meaning in Markdown’s formatting syntax. For example, if you wanted to surround a word with literal asterisks you can use backslashes before the asterisks, like this,

\*literal asterisks\*

because otherwise asterisks are use for italic fonts, like this: *literal asterisks*. 

\

### 2.6 Links

To turn a URL or an email address into a link, enclose it in angle brackets, 

<https://www.google.it/>

or you can enclose the link text in brackets and then follow it immediately with the URL in parentheses, like this:

My favorite search engine is [GOOGLE](https://www.google.it/)

\

### 2.7 Writing math

LaTex equations, inline, like this $\sqrt{\frac{1}{\pi}}$ or displayed like this $$f(x)=\lambda e^{-\lambda x}$$

A pair of tildes (\~) turn text to a subscript, e.g., H~3~PO~4~. A pair of carets (\^) produce a superscript, e.g., Cu^2+^.

\

### 2.8 Tables

To add a table, use three or more hyphens (---) to create each column’s header, and use pipes (\|) to separate each column. For compatibility, you should also add a pipe on either end of the row.

You can align text in the columns to the left, right, or center by adding a colon (:) to the left, right, or on both side of the hyphens within the header row.

| Example  |      Of       |  Table |
|:---------|:-------------:|-------:|
| col 1 is | left-aligned  | \$1600 |
| col 2 is |   centered    |   \$12 |
| col 3 is | right-aligned |    \$1 |

\

## 3. knitr

-   Is an engine for dynamic report generation with R. It is a package in the programming language R that enables integration of R code into `LaTeX`, LyX, HTML, `Markdown`, AsciiDoc, and reStructuredText documents.

-   The basic idea in knitr is that your regular text document will be interrupted by `chunks` of code delimited in a special way.

\

```{r}
paste("Hello", "World")
rnorm(1:5)
```

\

It is possible to add a chunk name inside braces after the `r`.

\

```{r my-first-chunk}
paste("Hello", "World")
rnorm(1:5)
```

\

The chunk name is not necessarily required however, it is good practice to give each chunk a unique name to support more advanced knitting approaches. Furthermore, in the RStudio IDE, you can navigate to specific chunks. Open the code chunk navigation window, located in the bottom left-hand side of the Source pane.

\

### 3.1 Some chunk options

If `echo=FALSE` knitr will not display the code in the final document, but the results will appear in the final document. This is a useful way to embed figures.

```{r try-echo-false, echo=FALSE}
paste("Hello", "World")
rnorm(1:5)
```

\

If `eval=FALSE` then knitr will not run the code chunk

```{r try-eval-false, eval=FALSE}
paste("Hello", "World")
rnorm(1:5)
```

\

If `include=FALSE` knitr will run the chunk code but not include the chunk in the final document. However, the results can be used by other chunks.

```{r try-include-false, include=FALSE}
paste("Hello", "World")
rnorm(1:5)
```

\

### 3.2 Inline R code 

Inline R code is embedded in the narratives of the document using the syntax  \` r \`, for example \` r sqrt(9)\*3\`  will result in the output `r sqrt(9)*3`. 

\

### 3.3 Figures

```{r include-figure}
hist(rnorm(1000), freq= FALSE, col="blue")
```

some options for plots

```{r figure-with-options, fig.height=3, fig.width=3, fig.align="right"}
hist(rnorm(1000), freq= FALSE, col="blue")
```

\

### 3.4 Tables

By default, R Markdown displays data frames and matrices as they would be in the R terminal.

```{r example-of-table}
library(ISLR2)
data(Auto)
Auto[1:5,]
```

\

If you prefer that data be displayed with additional formatting you can use the function `kable()` in the package `knitr`

```{r function-kable-for-talbles}
library(knitr)
kable(Auto[1:5,], format = "simple")
#
# "c"enter, "l"eft and "r"ight alignment of columns
#
kable(Auto[1:5,], format = "simple", align = "clrclrccl")
#
# change column names
#
kable(Auto[1:5,], format = "simple", align = "clrclrccl", col.names = LETTERS[1:9])
```

\

## 4 RStudio’s visual markdown editor

The RStudio IDE includes a visual markdown editor that displays changes in real-time and provides support for technical writing. 

To switch into the visual mode for a markdown document click the “Visual” button located on the left side of the editor toolbar.

Visual editing mode generates markdown using Pandoc, that is an open-source document converter. This means that in some cases your markdown will be rewritten to conform to standard Pandoc idioms. That is, **the Pandoc generated markdown that might differ from your own markdown writing style**.
