---
title: "Application of statistical analysis and modelling to Seoul Bike Sharing Demand dataset"
author: "Vasile Bezer (2124011), Second Student (12345678)"
font: 10pt
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    toc: false
    number_sections: true
  word_document:
    toc: false
    number_sections: true
---

\

# Introduction
A regression problem involves predicting a continuous numeric outcome using one or more input features. Linear regressors are based on the assumption that
there is a linear relationship between both the dependent variable and independent variables. This project focuses on the applications of linear regressors
on the Seoul Bike Sharing Demand dataset which incorporates real world bike sharing count of the city of Seoul with weather context ([the dataset can be retreived here](https://archive.ics.uci.edu/dataset/560/seoul+bike+sharing+demand)).
As major cities worldwide are moving towards smart mobility we hope to gain insight into how open-vehicle sharing systems could be optimized for efficiency 
through out the year by finding relationships between number of shares and weather.

The outcomes of the project are expected to be the following:

1. Understanding the siginficance of weather on bike sharing systems
2. Estimating the impact of predictors on the target variable
3. Building statistical models exploiting this dataset

\

# Dataset insight

We proceed our in depth look at the dataset by importing and looking at the summary. The format of the dataset is csv, to read it 
we use the built-in function read.csv of R.

```{r}
seoul.bike.data <- read.csv(
	file = "../data/SeoulBikeData.csv"
	,sep = ","
);
nrow(seoul.bike.data); ncol(seoul.bike.data); summary(seoul.bike.data);
```

We notice that the proposed dataset consists of 1 target, namely Rented Bike count, and 13 predictors whose characteristics are described by 8760 rows.
The description of each predictor is given by the dataset's source as follows:

| Variable Name         | Description                                          | Role    | Type        | Units | Missing Values |
| :-------------------- | :--------------------------------------------------- | :------ | :---------- | :---: | :------------- |
| Date                  | year-month-day                                       | Feature | Date        |       | no             |
| Rented Bike count     | Count of bikes rented at each hour                   | Feature | Integer     |       | no             |
| Hour                  | Hour of the day                                      | Feature | Integer     |       | no             |
| Temperature           | Temperature in Celsius                               | Feature | Continuous  |   C   | no             |
| Humidity              | %                                                    | Feature | Integer     |   %   | no             |
| Windspeed             | m/s                                                  | Feature | Continuous  |  m/s  | no             |
| Visibility            | 10m                                                  | Feature | Integer     |  10m  | no             |
| Dew point temperature | Celsius                                              | Feature | Continuous  |   C   | no             |
| Solar radiation       | MJ/m2                                                | Feature | Continuous  | Mj/m2 | no             |
| Rainfall              | mm                                                   | Feature | Integer     |  mm   | no             |
| Snowfall              | cm                                                   | Feature | Integer     |  cm   | no             |
| Seasons               | Winter, Spring, Summer, Autumn                       | Feature | Categorical |       | no             |
| Holiday               | Holiday/No holiday                                   | Feature | Binary      |       | no             |
| Functional Day        | NoFunc (Non-Functional Hours), Fun(Functional hours) | Feature | Binary      |       | no             |

## Missing value imputation

We notice that none of the predictors are reported as missing in the dataset's source, we proceed by checking this fact:

```{r}
ceiling( colSums(is.na(seoul.bike.data)) / nrow(seoul.bike.data) * 100 );
```

A key part of data preprocessing is dealing with missing values. These gaps can happen due to incomplete information or other reasons.
If not handled properly, they can result in incorrect predictions. It's important to find and fix any missing values but
in our case this dataset doesn't present this problem.

## Data preprocessing

Our dataset contains the following categorical variables: Seasons, Holiday, Functional Day. As seen before R treats them
as character variables but we need to we therefore proceed to transform them into factors for R to interpret them correctly.

```{r}
seoul.bike.data$Seasons 		<- factor( seoul.bike.data$Seasons );
seoul.bike.data$Holiday 		<- factor( seoul.bike.data$Holiday, ordered = TRUE );
seoul.bike.data$Functioning.Day <- factor( seoul.bike.data$Functioning.Day );
```

Transforming everything into factors servers the following purpose

We also need to transform our predictors Date and Hour into a date format correctly interpretable by R.

```{r}
seoul.bike.data$Date <- as.Date( seoul.bike.data$Date, format="%d/%m/%Y" );
```

Finally, we add the intercept to our data and transform the dataset into a matrix format.

```{r}
seoul.bike.datam <- cbind(1, seoul.bike.data);
seoul.bike.datam <- as.matrix( seoul.bike.data );
```

\

# Exploratory Data analysis

We begin this section of EDA by taking a look at the correlation matrix for our continuous predictors.


```{r}

```

## Correlation matrix

```{r}
library(corrplot, warn.conflicts = FALSE); corrplot( 
	cor(seoul.bike.data[2:11])
	,method = "number"
	,diag = FALSE
	,tl.cex = 0.8
	,number.cex = 0.6
	,tl.col = "black" 
	);
```

## Partial correlation graph

Partial correlation allows us to investigate the presence of linear relationships between variables with the effect of a set of controlling random
variables removed. This in turn allows us to capture relationships more in depth compared to the classic correlation matrix. We proceed to build this graph
by using the treshold method.

```{r}
library(igraph, warn.conflicts = FALSE); library(gRbase, warn.conflicts = FALSE);

threshold <- 1/;
variance.matrix <- var( seoul.bike.data[2:11], use="pairwise.complete.obs" );
covariance.matrix <- -cov2cor( solve(variance.matrix) );
independence.matrix <- abs( covariance.matrix ) > threshold;
diag(independence.matrix) <- 0; # remove self dependencies
independence.graph <- as( independence.matrix, "igraph" );
plot(independence.graph);
```

We can notice how the graph clearly presents us with the fact that our target is correlated with a multitude of predictors in our dataset.
On the contrary, Snowfall and Rainfall seem to not be linaerly correlated to the target but this fact alone does not mean we should remove 
these variables from our analysis.

## Scatter plots

## Distribution

We proceed with the representation of scatter plots to visually identifiyif we can see
evident relationships between our predictors and our target, in our case the target is Rented Bike count.

```{r}
plot(seoul.bike.data$Rented.Bike.Count, seoul.bike.data$Temperature, pch = 20);
plot(seoul.bike.data$Rented.Bike.Count, seoul.bike.data$Hour, pch = 20);
```

## Distribution



# Modelling

## Diagnostics

## Residual analysis

## Outliers and high-leverage points

## Collinearity

## ANOVA



# Conclusions


